Higher-Order unification for free!

### 2024-04-15 paper/main.tex
section 8

In this paper we show a minimized example. The full code is there.
But we also have to code things in Coq-Elpi.

The main difference between the presentation in the previous sections and
the actual implementation for Coq is that the main loop \hrun is replaced by
the one of Prolog that chains calls to the unification procedure. In order
implement the store of links we resort to Elpi's CLP engine and
use constraints (suspended goals) to represent links, and constraint
handling rules to implement progress operations involving more than one link.

\todo{finish}
about the progress of 1 link:

\begin{elpicode}
link-eta L R :- suspend-condition L R Holes, !,
  declare_constraint (link-eta L R) Holes.
link-eta L R :-
  progress. % e.g. L = R.
\end{elpicode}

about the progress of 2 links:

@@

Remark how the invariant about uvar arity makes this easy, since LX1 and LX2
have the same length. Also note that N1 only contains the names of the first
link (while relocate runs in the disjoint union) and Elpi ensures that
T2' can live in N1.

>>>>

In this paper we did study a compiler and simulation loop on a minimal language.
The actual implementation uses the Coq-Elpi meta language to both compile
the sequence of problems (the rules) and execute them, that is Elpi
plays the role of \Ho{} as well.

The main difference is that we cannot implement \hrun since it comes as part of
the runtime of the programming language. In particular the runtime iterates
\Ue, but \hstep also needs to check for \progress. Luckily Elpi extends~\cite{TASSI_2019}
$\lambda$Prolog with syntactic constraints (suspended goals) and
Constraint Handling Rules (CHR)~\cite{fruehwirth2017constraint} to operate on them.
An important design choice is that constraints are suspended on a user
computed list of unification variables, and are resumed as soon as one of them
is assigned, before any other existing goal is considered. In turn this
property enables a timely activation of links that grants~\fullref{prop:fidelity}.
For brevity we only provide two pseaudo-code snippets. The first one depicts
how \linketa are suspended or make \progress.

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{improve code}

\noindent
The second snippet illustrate the deduplication of \linketa. The syntax
<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
operator) \elpi{N} heigenvariables. Sequents are presented to
CHR with their higher order unification variables replaced by
``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
for a fresh constant $f_A$. Frozen constants are ``defrost'' when
they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).

@@

\noindent
The first directive matches a constraint
whose \lhs is a variable \elpiIn{X}; the second mathes \emph{and removes}
a constrain on the same variable; the third is a test that accessorily
relocates \elpiIn{T2} and the latter crafs the new goal that unifies
\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
heigenvariables \elpiIn{N1}.

% Remark that \fullref{inv:uvaarity} makes it possible to relocate \elpiIn{T2}
% using the name-to-name mapping obtained by zipping
% the lists (of equal length) \elpiIn{LX1} and \elpiIn{LX2}.
